---
title: 'Machine Learning: Exercises 1'
author: "Your Name"
date: "Updated `r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
### Install packages if necessary
if (!require("pacman")) install.packages("pacman") # package for loading and checking packages :)
pacman::p_load(tidyverse, magrittr, 
               tidymodels,
               tidytext
               )
```

# Introduction

Its time tnow to show what you can do and practiced what you learned. Your task will be the prediction of wine quality according to its chemical composition. You will master the following tasks:

1. Apply some text manipulation skills on new data.
2. Develop an predictive model based on text data.
# The Data

```{r}
data <- read_csv(url("https://www.ocf.berkeley.edu/~janastas/trump-tweet-data.csv")) %>% as_tibble() 
```

```{r}
data %>% glimpse()
```

```{r}
data %>% head()
```

Lets define our outcome variable. We would like to predict if a tweet went viral. Lets construct a variable.

```{r}
data %<>%
  mutate(y = (Retweets > 63) %>% as.factor() )
```

# Your turn

Check out the tweets, is there something we could easily do to clean up the tweets?

```{r}

```

Lets analyze a bit, do we see differences in words used between viral and non-viral tweets?

```{r}

```

Ok, can we develop a predictive model thereof?

```{r}

```



